NFL Draft Analysis

### 2020-05-13 draft_analysis.Rmd
Updated tutorial and added some supporting links

The result will be stored in the dataframe *master_df* with entities for each player.

Note: We will also do some process in this step as well, but we will go into more detail on that when we repeat the processing steps in another dataframe later in the walkthrough.

@@

Load the NFL draft history data from *master_df*, containing the draft data from rounds 1-7 from the 1970-2016 NFL draft. We will only keep the observations that do not have any missing data entries. Also, we will convert the necessary columns to integers and factors (rather than chars). This way, we can use the columns values to group and to produce regression lines in our data visualization steps later.

@@

Now let's process our 2020 Draft data. First, we will get rid of columns we don't care about, then remove blank rows, add year drafted to each observation, and convert the necessary columns to integers and factors, all like we did for the historical draft dataset. We will also filter this dataframe to only include quarterbacks, wide receivers, and running backs.

@@

# get rid of blank rows
df_2020 <-df_2020[!(df_2020$Player=="Player"),]

@@

Now let's create a dataframe in which we can compare different positions that were drafted. First, we want to break up our data into 4 periods. Then for each period, we will take average career value of each position grouped by each round that the player was selected. We will make a new *pos_df* dataframe with these groupings.

@@

Let's also add a dataframe *off_df* that is the same as *master_df*, but contains a variable **roty** that is Yes when the player that was drafted from 1970-2016 was Offensive Rookie of the Year, and No otherwise. We will also filter the dataframe to only contain offensive skill position players, like we did for the 2020 draft data.

@@

To do this, we will plot the average value for all the players for each year since 1970, grouped by round.

@@

As we would expect the trend is that there is a decline in value until about pick 100, where the career value seems to be about the same from about pick 100-250. This would suggest that seelcting a QB at picks 100-150 would, statistically speaking, provide less bang for your buck, as you could likely get a player of the same quality with some of your later draft capital.

@@

@@

First, we will set the predictor variables to **Tm**, **Pos**, and **Pick**, and the sole outcome variable as **roty**. Then we will use the *train()* method from the caret package to train the model using the *off_df* variable that contains all our historical draft data on all the offensive skill position players.

@@

Let's test our model against the 2017, 2018, and 2019 datasets. To do this, we will use the *predict()* function in the caret package and feed it the test data. From there,  *predict()* will calculate probabilities that the model is sure about it choice that each observation from *test_df* will fall under the "Yes" and "No" outcome categories. So the **Yes** variable corresponds to the probability the model is sure about its choice that the player will become Rookie MVP, and **No** = probability the model is sure the player will not be the Rookie MVP.

@@

Now let's get the top 10 players that the model is most certain will be the Offensive Rookie of the Year for 2017, 2018, and 2019.

@@

Now let's get the top 10 players that the model is most certain will be the Offensive Rookie of the Year

>>>>

To scrape the data, we will use the *read_html()*, *html_node()*, and *html_table()* methods in the **rvest** package. More information on **rvest** is available here: https://blog.rstudio.com/2014/11/24/rvest-easy-web-scraping-with-r/

First, we store the html file from our draft data website into the *draft_html* variable. From there, we search for the first occurence of the "table" element using *html_node()*, and then we parse the HTML data results into a dataframe using the *html_table()*. After parsing, we see that the column names are in the first row of the data, so we set the column names to be the first row of data with *names(df) <- as.character(unlist(df[1,]))*. Then we delete the first row of data (that still contains the column names) so all of the rows just contain player data.

Since each years draft data is on a different website, we loop through all of the website, and after some more processing (see note), we use *rbind()* to add each dataframe vertically to our *master_df* dataframe.

The result will be all of the 1970-2016 draft data stored in the dataframe *master_df* with entities for each player.

*Note: We will also do some process in this step as well, but we will go into more detail on that when we repeat the processing steps in another dataframe later in the walkthrough.

@@

Load the NFL draft history data from *master_df*, containing the draft data from rounds 1-7 from the 1970-2016 NFL draft. The first line in this chunk will ensure that we will only keep the observations that do not have any missing data entries. Then, we will convert the necessary columns to integers and factors (rather than chars). This way, we can use the columns values to group and to produce regression lines in our data visualization steps later.

@@

Now let's process our 2020 Draft data. We will be using the *filter()* and *mutate()* methods found in the **dplyr** package in the **tidyverse** package. More info on **dplyr** can be found here: https://dplyr.tidyverse.org/

First, we will get rid of columns we don't care about. 
Next, we only keep rounds 1-7 using *filter()* with the conditional that **Rnd** variable must be less than 8. 
There seem to be seperator rows between rounds that contain the column names again (probably for readability on the website), so we get rid of them by choosing to keep only rows where the **Player** variable does not equal "Player". 
We then add year drafted to each observation using *mutate()*, which allows us to add variables to an existing dataframe, and again use *filter()* to only include quarterbacks, wide receivers, and running backs. 
We noticed that some of the age data is missing for the players, so we fill in all of the missing age values with 22, since that is the most common age for a player to enter the pros.
We convert the necessary columns to integers and factors, all like we did for the historical draft dataset.

@@

# get rid of seperator rows
df_2020 <-df_2020[!(df_2020$Player=="Player"),]

@@

Now let's create a dataframe in which we can compare different positions that were drafted. First, we want to break up our data into 4 periods. We accomplish this with the *cut()* method on our **year** variable with parameter 'breaks' = 4. This will break up all of the data into 4 periods based on year, more information on cut is available here: https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/cut

We will use *mutate()* again to add a variable **period** for which period this entry is in.
Then for each period, we will take average career value of each position using *summarise()*, grouped by each round, period and position that the player was selected using *group_by()*. Both of these methods are also available in the **dplyr** package.

We will make a new *pos_df* dataframe with these groupings.

@@

Let's also add a dataframe *off_df* that is the same as *master_df*, but contains a variable **roty** that is Yes when the player that was drafted from 1970-2016 was Offensive Rookie of the Year, and No otherwise. Again, since we are adding a new variable, we use *mutate()*, with a call to *ifelse()*. The condition in the ifelse will check if the entry's player name exists in the **Player** column of our rookie of the year table, if it exists it returns "Yes", and "No" otherwise. Lastly, we convert this new variable into a factor w/ 2 levels.

We will also filter the dataframe to only contain offensive skill position players, like we did for the 2020 draft data.

@@

To do this, we will plot the average value for all the players for each year since 1970, grouped by round. We will create our plots using **ggplot2** which is available in the **tidyverse** package, more info can be found here: https://ggplot2.tidyverse.org/

@@

As we would expect the trend is that there is a decline in value until about pick 100, where the career value seems to be about the same from about pick 100-250. So the data suggests that selecting a QB at picks 100-150 would provide less bang for your buck, as you could likely get a player of the same quality with some of your later draft capital.

@@

All of our machine learning goodness uses methods from the **caret** package, more info on this package can be found here: https://cran.r-project.org/web/packages/caret/vignettes/caret.html

@@

First, we will set the predictor variables to **Tm**, **Pos**, and **Pick**, and the sole outcome variable as **roty**. Then we will use the *train()* method from the **caret** package to train the model using the *off_df* variable that contains all our historical draft data on all the offensive skill position players. We tell *train()* what predictor and outcome variables to use, as well as to use the gbm method with the resampling control that we set previously.

@@

Let's test our model against the 2017, 2018, and 2019 datasets. To do this, we will use the *predict()* function in the **caret** package and feed it the test data. Since we set parameter 'type' to "prob",  *predict()* will calculate probabilities that the model is sure about it choice that each observation from *test_df* will fall under the "Yes" and "No" outcome categories. So the **Yes** variable corresponds to the probability the model is sure about its choice that the player will become Rookie MVP, and **No** = probability the model is sure the player will not be the Rookie MVP.

@@

Now let's get the top 10 players that the model is most certain will be the Offensive Rookie of the Year for 2017, 2018, and 2019 respectively.

@@

Now let's get the top 10 players that the model is most certain will be the Offensive Rookie of the Year in 2020.
